Signatures and Induction Principles for Higher Inductive-Inductive Types

### 2019-01-31 paper.tex
Fix latexmk errors

\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\REN}{\mathsf{REN}}
\newcommand{\op}{\mathsf{op}}
\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\PSh}{\mathsf{PSh}}
\newcommand{\FamPSh}{\mathsf{FamPSh}}
\renewcommand{\ll}{\llbracket}
\providecommand{\rr}{\rrbracket}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\R}{\mathsf{R}}
\newcommand{\TM}{\mathsf{TM}}
\newcommand{\NE}{\mathsf{NE}}
\newcommand{\NF}{\mathsf{NF}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\renewcommand{\u}{\mathsf{u}}
\renewcommand{\ne}{\mathsf{ne}}
\newcommand{\nf}{\mathsf{nf}}
\newcommand{\lQ}{\mathsf{lQ}}
\newcommand{\lU}{\mathsf{lU}}
\renewcommand{\lq}{\mathsf{lq}}
\newcommand{\lu}{\mathsf{lu}}
\newcommand{\cul}{\ulcorner}
\newcommand{\cur}{\urcorner}
\newcommand{\norm}{\mathsf{norm}}
\newcommand{\Nf}{\mathsf{Nf}}
\newcommand{\Ne}{\mathsf{Ne}}
\newcommand{\Nfs}{\mathsf{Nfs}}
\newcommand{\Nes}{\mathsf{Nes}}
\newcommand{\ID}{\mathsf{ID}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\nat}{\,\dot{\rightarrow}\,}
\newcommand{\Nat}{\mathsf{Nat}}
%% \renewcommand{\S}{\overset{\mathsf{s}}{\ra}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\tr}[2]{\ensuremath{{}_{#1 *}\mathopen{}{#2}\mathclose{}}}
\newcommand{\C}{\mathsf{C}}
\newcommand{\Code}{\mathsf{Code}}

@@

\newcommand{\Elim}{\mathsf{Elim}}
\newcommand{\elim}{\mathsf{elim}}
\newcommand{\Rec}{\mathsf{Rec}}
\newcommand{\record}{\mathsf{record}}
\newcommand{\funext}{\mathsf{funext}} \newcommand{\Q}{\mathsf{Q}}
\newcommand{\T}{\mathsf{T}} \newcommand{\leaf}{\mathsf{leaf}}
\newcommand{\node}{\mathsf{node}} \newcommand{\perm}{\mathsf{perm}}
\newcommand{\coe}{\mathsf{coe}} \newcommand{\vz}{\mathsf{vz}}
\newcommand{\vs}{\mathsf{vs}} \newcommand{\untr}{\mathsf{untr}}
\newcommand{\from}{\mathsf{from}}
\newcommand{\fromeq}{{\mathsf{from}\hspace{-0.3em}\equiv}}
\newcommand{\fromsimeq}{{\mathsf{from}\hspace{-0.3em}\simeq}}
\newcommand{\Model}{\mathsf{Model}}
\newcommand{\DModel}{\mathsf{DModel}}
\newcommand{\module}{\mathsf{module}}
\newcommand{\open}{\mathsf{open}}
\renewcommand{\P}{\mathsf{P}} \newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}} \newcommand{\false}{\mathsf{false}}
\renewcommand{\not}{\mathsf{not}} \newcommand{\0}{\mathsf{0}}
\newcommand{\1}{\mathsf{1}} \renewcommand{\Pr}{\mathsf{Pr}}
\newcommand{\PrNat}{\mathsf{PrNat}} \newcommand{\J}{\mathsf{J}}
\newcommand{\wkV}{\mathsf{wkV}} \renewcommand{\r}[1]{{\P_{#1}}}
\newcommand{\stab}{\mathsf{stab}} \newcommand{\NTy}{\mathsf{NTy}}
\newcommand{\isDec}{\mathsf{isDec}} \newcommand{\dec}{\mathsf{dec}}
\newcommand{\yes}{\mathsf{yes}} \newcommand{\no}{\mathsf{no}}
\newcommand{\case}{\mathsf{case}} \newcommand{\inj}{\mathsf{inj}}
\newcommand{\K}{\mathsf{K}} \newcommand{\lb}{\langle}
\newcommand{\rb}{\rangle} \newcommand{\Decl}{\mathsf{Decl}}
\newcommand{\Core}{\mathsf{Core}} \newcommand{\IND}{\mathsf{ind}}
\newcommand{\Id}{\mathsf{Id}} \newcommand{\Base}{\mathsf{Base}}
\newcommand{\transport}{\mathsf{transport}}
\newcommand{\I}{\mathsf{I}} \newcommand{\E}{\mathsf{S}}
\newcommand{\transp}{\mathsf{transp}}
\newcommand{\Transp}{\mathsf{Transp}}
\newcommand{\W}{\mathsf{W}}
\newcommand{\Fin}{\mathsf{Fin}} \newcommand{\fzero}{\mathsf{fzero}}
\newcommand{\fsuc}{\mathsf{fsuc}}
\newcommand{\inv}{\mathsf{inv}}
\newcommand{\con}{\mathsf{con}}
\newcommand{\LEFT}{\mathsf{left}}
\newcommand{\RIGHT}{\mathsf{right}}
\newcommand{\seg}{\mathsf{seg}}
\newcommand{\Int}{\mathsf{Int}}
\renewcommand{\in}{\mathbin{\hat:}}
\renewcommand{\hat}[1]{{\color{BrickRed}{#1}}}
\newcommand{\blc}[1]{{\color{Black}{#1}}}
\newcommand{\vdashh}{\mathbin{\hat\vdash}}
\newcommand{\rah}{\mathbin{\hat\ra}}
\newcommand{\commah}{\hat,\,}
\newcommand{\timesh}{\mathbin{\hat\times}}
\newcommand{\eqh}{\mathbin{\hat=}}
\newcommand{\TR}{\hat{\mathsf{tr}}}
\newcommand{\ap}{\hat{\mathsf{ap}}}
\newcommand{\apd}{\hat{\mathsf{apd}}}
\renewcommand{\tt}{\hat{\mathsf{tt}}}
\newcommand{\Tel}{\hat{\mathsf{Tel}}}
\newcommand{\Type}{\hat{\mathsf{Type}}}
\newcommand{\emptytel}{\hat{\epsilon}}
\newcommand{\telext}{\mathbin{\hat{\lhd}}}
\newcommand{\ltel}{\hat{\ll}}
\newcommand{\rtel}{\hat{\rr}}
\newcommand\pp{\ensuremath{\hat{\mathbin{+\mkern-10mu+}}}}
\newcommand{\semicol}{\hat;\,}
\newcommand{\targetass}{\hat{\Gamma}\semicol}
\newcommand{\Alg}{\mathsf{Alg}}
\newcommand{\DisplayedAlg}{\mathsf{DisplayedAlg}}
\newcommand{\Morphism}{\mathsf{Morphism}}
\newcommand{\Section}{\mathsf{Section}}
\newcommand{\ass}{\mathsf{ass}}
\newcommand{\idl}{\mathsf{idl}}
\newcommand{\idr}{\mathsf{idr}}
\newcommand{\inv}{\mathsf{inv}}
\newcommand{\ID}{\mathsf{\ID}}

@@

(4) $\U$ is closed under the equality type, with eliminator $\J$ and a weak
(propositional) $\beta$-rule. Weakness is required because the translations in
Sections \ref{sec:morphisms} and \ref{sec:sections} do not preserve this
$\beta$-rule strictly. We explain this in more detail in Sections TODO and
TODO. Adding equality to the theory of codes allows higher constructors and
inductive equality parameters as well. We can now define the higher-inductive
circle as the following context:
\[
\boldsymbol{\cdot},\,\,\,S^1:\U,\,\,\,base:\underline{S^1},\,\,\,loop:\underline{base =_{S^1} base}
\]

@@

\begingroup
\allowdisplaybreaks
\begin{alignat*}{5}
  & \boldsymbol{\cdot}^\A && :\equiv \hat{\top} \\
  & (\Delta,x:A)^\A && :\equiv \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\timesh A^\A\,\hat{\gamma} \\
  & x^\A\,\hat{\gamma} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma} \\ % TODO: write De Bruijn versions
  & \U^\A\,\hat{\gamma} && :\equiv \Type_{\hat{0}} \\
  & (\underline{a})^\A\,\hat{\gamma} && :\equiv a^\A\,\hat{\gamma} \\
  & ((x:a)\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)} \\
  & (t\,u)^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\A\,\hat{\gamma}\hat{)} \\
  & ((\hat{x}\in \hat{A})\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A})\rah  B^\A\,\hat{\gamma} \\
  & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u} \\
  & (t=_a u)^\A\,\hat{\gamma} && :\equiv t^\A\,\hat{\gamma} \eqh u^\A\,\hat{\gamma} \\
  & \refl^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\A\,\hat{\gamma} && :\equiv \hat{\J}_{\hat{(}a^\A\,\hat{\gamma}\hat{)}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}\hat{\lambda} \hat{x}\,\hat{z}.p^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\commah\hat{z}\hat{)}\hat{)}}\,\hat{(}pr^\A\,\hat{\gamma}\hat{)}\,_{\hat{(}u^\A\,\hat{\gamma}\hat{)}}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)} \\
  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
  & ((\hat{x}\in \hat{A})\ra b)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^\A\,\hat{\gamma} \\
  & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u}
\end{alignat*}
\endgroup

@@

However, out current approach does not scale to the point where we have to worry
about higher categories. We explain in the following. The natural next step
towards a categorical semantics would be defining a $\blank^\ID$ translation,
which computes identity homomorphisms:

\[
\infer{\hat{\Gamma}\vdashh\Delta^\ID\,\hat{:\,(\gamma : \blc{\Delta^\A})\ra \blc{\Delta^\M}\,\gamma\,\gamma}}
      {\hat{\Gamma}\vdash\Delta}
\hspace{1em}
\infer{
  \hat{\Gamma}\vdashh A^\ID\,\hat{:\,(\gamma : \blc{\Delta^\A})(t : \blc{A^\A}\,\gamma)\ra \blc{A^\M}\,\gamma\,\gamma\,(\blc{\Delta^\ID}\,\gamma)\,t\,t }}
      {\hat{\Gamma}\semicol\Delta\vdash A}
\]

\[
\infer{
  \hat{\Gamma}\vdashh t^\ID\,\hat{:\, (\gamma : \blc{\Delta^\A})\ra
       \blc{t^\M}\,\gamma\,\gamma\,(\blc{\Delta^\ID}\,\gamma) = \blc{A^\ID}\,\gamma\,(\blc{t^\A}\,\gamma) }}
      {\hat{\Gamma}\semicol\Delta\vdash t : A}
\]

Here, the interpretation of terms witnesses functoriality: $t^\M$ maps
identity morphisms in $\Delta$ to displayed identity morphisms in $A$.
We translate $\U$ to identity functions:
\[
\U^\ID\,\hat{\gamma\,A} :\equiv \hat{\lambda\,(x : A).\, x}
\]
Note that $\U$ and $\U[\sigma]$ are definitionally equal in the theory of
signatures, where $\sigma$ is a parallel substitution. The translation of the
latter would be the following (omitting many details, including the handling of
substitutions in the translation):
\[
(\U[\sigma])^\ID\,\hat{\gamma\,A} \equiv \hat{\TR_{(\lambda\,x. A\ra A)}\,(\sigma^\ID\,\gamma)\,(\lambda\,x.\, x)}
\]
Here, $\hat{\sigma^\ID\,\gamma}$ yields an equation for functoriality of
$\hat{\sigma}$. Since the transport is constant, the result is propositionally
equal to $\hat{\lambda\,x.\, x}$. However, that is not enough, since we need to
preserve $\U = \U[\sigma]$ up to definitional equality.

There are numerous unavoidable instances of such failures of strict preservation
of definitional equalities for $\blank^\ID$, and they similarly arise when we
try to construct composition of homomorphisms. We consider three potential
solutions for this strictness problem:
\begin{enumerate}
\item Solving the coherence problem. This would allow us to interpret signatures
      into the metatheory, allowing preservation of definitional equality up
      to propositional metatheoretic equality.
\item Reformulating the syntax of signatures so that definitional equalities
      become weak propositional equalities. We already use weak $\beta$ for $\J$,
      can we do so elsewhere? However, this would result in an unusual and very
      inconvenient syntax of signatures, because we would need to weaken even
      basic substitution rules.
\item Instead of interpreting signatures into an UIP-free type theory, we
      interpret them into classical combinatorial structures for higher groupoids
      and categories, e.g.\ into simplicial sets. The drawback is that now we do not
      have the convenient synthetic notion of higher structures, provided by the syntax
      of type theory, and instead we have to manually build up these
      structures. Needless to say, this makes machine-checked formalization much
      more difficult. We have found mechanized formalization invaluable for the
      current paper, and it would be painful to abandon it in further research.
\end{enumerate}

@@

%% Nevertheless, the current approach of using syntactic translations is a sweet
%% spot for computing induction principles in a relatively simple way.

@@

There are three additional development artifacts to the current work: a Haskell
implementation, a shallow Agda formalization and a deep Agda formalization. All
three are available from the homepage of the first author.

The Haskell implementation takes as input a file which contains a
representation of a $\hat{\Gamma}\vdash\Delta$ specification of a
HIIT. Then, it checks the input with respect to the rules in Figure
\ref{sigrules}, and outputs an Agda-checkable file which contains the
results of the $\blank^\C$, $\blank^\M$ and $\blank^\E$
translations. It comes with examples, including the ones in this
paper, indexed W-types \cite{morris09indexed}, the dense completion
\cite[Appendix A.1.3]{forsberg-phd} and several HITs from
\cite{HoTTbook} including the definition of Cauchy reals. It can be
checked that our implementation computes the expected elimination
principles in these cases.

The shallow Agda formalization embeds both the source and target theories
shallowly into Agda: it represents types as Agda types, functions as
Agda functions, and so on. We also leave the $\blank^\C$ operation
implicit. We state each case of the $\blank^\M$ and $\blank^\E$
translations as Agda functions from all induction hypotheses to the
result type of the translation, which lets us ``typecheck'' the
translation. We have found that this style of formalization is
conveniently light, but remains detailed enough to be useful. We also
generated some of the code of the Haskell implementation from this
formalization.

The deep Agda formalization still uses a shallow embedding of the
target type theory, but it uses a deep embedding of the source theory,
in the style of \cite{ttintt}. In the formalization we merge the three
translations into a single model construction. This allows us to prove
strict preservation of definitional equalities in the substitution
calculus of the source theory, in contrast to the shallow
formalization, where we cannot reason about definitional equalities of
Agda terms. Due to technical challenges, this formalization uses
transport instead of $\J$ in the source theory, but this still covers
a rather large class of HIIT definitions.

>>>>

\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\REN}{\mathsf{REN}}
\newcommand{\op}{\mathsf{op}}
\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\PSh}{\mathsf{PSh}}
\newcommand{\FamPSh}{\mathsf{FamPSh}}
\renewcommand{\ll}{\llbracket}
\providecommand{\rr}{\rrbracket}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\R}{\mathsf{R}}
\newcommand{\TM}{\mathsf{TM}}
\newcommand{\NE}{\mathsf{NE}}
\newcommand{\NF}{\mathsf{NF}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\renewcommand{\u}{\mathsf{u}}
\renewcommand{\ne}{\mathsf{ne}}
\newcommand{\nf}{\mathsf{nf}}
\newcommand{\lQ}{\mathsf{lQ}}
\newcommand{\lU}{\mathsf{lU}}
\renewcommand{\lq}{\mathsf{lq}}
\newcommand{\lu}{\mathsf{lu}}
\newcommand{\cul}{\ulcorner}
\newcommand{\cur}{\urcorner}
\newcommand{\norm}{\mathsf{norm}}
\newcommand{\Nf}{\mathsf{Nf}}
\newcommand{\Ne}{\mathsf{Ne}}
\newcommand{\Nfs}{\mathsf{Nfs}}
\newcommand{\Nes}{\mathsf{Nes}}
%% \newcommand{\ID}{\mathsf{ID}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\nat}{\,\dot{\rightarrow}\,}
\newcommand{\Nat}{\mathsf{Nat}}
%% \renewcommand{\S}{\overset{\mathsf{s}}{\ra}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\tr}[2]{\ensuremath{{}_{#1 *}\mathopen{}{#2}\mathclose{}}}
\newcommand{\C}{\mathsf{C}}
\newcommand{\Code}{\mathsf{Code}}

@@

\newcommand{\Elim}{\mathsf{Elim}}
\newcommand{\elim}{\mathsf{elim}}
\newcommand{\Rec}{\mathsf{Rec}}
\newcommand{\record}{\mathsf{record}}
\newcommand{\funext}{\mathsf{funext}} \newcommand{\Q}{\mathsf{Q}}
\newcommand{\T}{\mathsf{T}} \newcommand{\leaf}{\mathsf{leaf}}
\newcommand{\node}{\mathsf{node}} \newcommand{\perm}{\mathsf{perm}}
\newcommand{\coe}{\mathsf{coe}} \newcommand{\vz}{\mathsf{vz}}
\newcommand{\vs}{\mathsf{vs}} \newcommand{\untr}{\mathsf{untr}}
\newcommand{\from}{\mathsf{from}}
\newcommand{\fromeq}{{\mathsf{from}\hspace{-0.3em}\equiv}}
\newcommand{\fromsimeq}{{\mathsf{from}\hspace{-0.3em}\simeq}}
\newcommand{\Model}{\mathsf{Model}}
\newcommand{\DModel}{\mathsf{DModel}}
\newcommand{\module}{\mathsf{module}}
\newcommand{\open}{\mathsf{open}}
\renewcommand{\P}{\mathsf{P}} \newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}} \newcommand{\false}{\mathsf{false}}
\renewcommand{\not}{\mathsf{not}} \newcommand{\0}{\mathsf{0}}
\newcommand{\1}{\mathsf{1}} \renewcommand{\Pr}{\mathsf{Pr}}
\newcommand{\PrNat}{\mathsf{PrNat}} \newcommand{\J}{\mathsf{J}}
\newcommand{\wkV}{\mathsf{wkV}} \renewcommand{\r}[1]{{\P_{#1}}}
\newcommand{\stab}{\mathsf{stab}} \newcommand{\NTy}{\mathsf{NTy}}
\newcommand{\isDec}{\mathsf{isDec}} \newcommand{\dec}{\mathsf{dec}}
\newcommand{\yes}{\mathsf{yes}} \newcommand{\no}{\mathsf{no}}
\newcommand{\case}{\mathsf{case}} \newcommand{\inj}{\mathsf{inj}}
\newcommand{\K}{\mathsf{K}} \newcommand{\lb}{\langle}
\newcommand{\rb}{\rangle} \newcommand{\Decl}{\mathsf{Decl}}
\newcommand{\Core}{\mathsf{Core}} \newcommand{\IND}{\mathsf{ind}}
\newcommand{\Id}{\mathsf{Id}} \newcommand{\Base}{\mathsf{Base}}
\newcommand{\transport}{\mathsf{transport}}
\newcommand{\I}{\mathsf{I}} \newcommand{\E}{\mathsf{S}}
\newcommand{\transp}{\mathsf{transp}}
\newcommand{\Transp}{\mathsf{Transp}}
\newcommand{\W}{\mathsf{W}}
\newcommand{\Fin}{\mathsf{Fin}} \newcommand{\fzero}{\mathsf{fzero}}
\newcommand{\fsuc}{\mathsf{fsuc}}
\newcommand{\inv}{\mathsf{inv}}
\newcommand{\con}{\mathsf{con}}
\newcommand{\LEFT}{\mathsf{left}}
\newcommand{\RIGHT}{\mathsf{right}}
\newcommand{\seg}{\mathsf{seg}}
\newcommand{\Int}{\mathsf{Int}}
\renewcommand{\in}{\mathbin{\hat:}}
\renewcommand{\hat}[1]{{\color{BrickRed}{#1}}}
\newcommand{\blc}[1]{{\color{Black}{#1}}}
\newcommand{\vdashh}{\mathbin{\hat\vdash}}
\newcommand{\rah}{\mathbin{\hat\ra}}
\newcommand{\commah}{\hat,\,}
\newcommand{\timesh}{\mathbin{\hat\times}}
\newcommand{\eqh}{\mathbin{\hat=}}
\newcommand{\TR}{\hat{\mathsf{tr}}}
\newcommand{\ap}{\hat{\mathsf{ap}}}
\newcommand{\apd}{\hat{\mathsf{apd}}}
\renewcommand{\tt}{\hat{\mathsf{tt}}}
\newcommand{\Tel}{\hat{\mathsf{Tel}}}
\newcommand{\Type}{\hat{\mathsf{Type}}}
\newcommand{\emptytel}{\hat{\epsilon}}
\newcommand{\telext}{\mathbin{\hat{\lhd}}}
\newcommand{\ltel}{\hat{\ll}}
\newcommand{\rtel}{\hat{\rr}}
\newcommand\pp{\ensuremath{\hat{\mathbin{+\mkern-10mu+}}}}
\newcommand{\semicol}{\hat;\,}
\newcommand{\targetass}{\hat{\Gamma}\semicol}
\newcommand{\Alg}{\mathsf{Alg}}
\newcommand{\DisplayedAlg}{\mathsf{DisplayedAlg}}
\newcommand{\Morphism}{\mathsf{Morphism}}
\newcommand{\Section}{\mathsf{Section}}
\newcommand{\ass}{\mathsf{ass}}
\newcommand{\idl}{\mathsf{idl}}
\newcommand{\idr}{\mathsf{idr}}
\renewcommand{\inv}{\mathsf{inv}}

@@

(4) $\U$ is closed under the equality type, with eliminator $\J$ and a weak
(propositional) $\beta$-rule. Weakness is required because the translations in
Sections \ref{sec:morphisms} and \ref{sec:sections} do not preserve this
$\beta$-rule strictly. We explain this in more detail in Sections
\ref{sec:coherence} and \ref{sec:categorical}. Adding equality to the theory of
codes allows higher constructors and inductive equality parameters as well. We
can now define the higher-inductive circle as the following context:
\[
\boldsymbol{\cdot},\,\,\,S^1:\U,\,\,\,base:\underline{S^1},\,\,\,loop:\underline{base =_{S^1} base}
\]

@@

\begingroup
\allowdisplaybreaks
\begin{alignat*}{5}
  & \boldsymbol{\cdot}^\A && :\equiv \hat{\top} \\
  & (\Delta,x:A)^\A && :\equiv \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\timesh A^\A\,\hat{\gamma} \\
  & x^\A\,\hat{\gamma} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma} \\
  & \U^\A\,\hat{\gamma} && :\equiv \Type_{\hat{0}} \\
  & (\underline{a})^\A\,\hat{\gamma} && :\equiv a^\A\,\hat{\gamma} \\
  & ((x:a)\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)} \\
  & (t\,u)^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\A\,\hat{\gamma}\hat{)} \\
  & ((\hat{x}\in \hat{A})\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A})\rah  B^\A\,\hat{\gamma} \\
  & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u} \\
  & (t=_a u)^\A\,\hat{\gamma} && :\equiv t^\A\,\hat{\gamma} \eqh u^\A\,\hat{\gamma} \\
  & \refl^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\A\,\hat{\gamma} && :\equiv \hat{\J}_{\hat{(}a^\A\,\hat{\gamma}\hat{)}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}\hat{\lambda} \hat{x}\,\hat{z}.p^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\commah\hat{z}\hat{)}\hat{)}}\,\hat{(}pr^\A\,\hat{\gamma}\hat{)}\,_{\hat{(}u^\A\,\hat{\gamma}\hat{)}}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)} \\
  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
  & ((\hat{x}\in \hat{A})\ra b)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^\A\,\hat{\gamma} \\
  & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u}
\end{alignat*}
\endgroup

@@

However, out current approach does not scale to the point where we have to worry
about higher categories. We explain in the following. The natural next step
towards a categorical semantics would be defining a $\blank^{\mathsf{ID}}$ translation,
which computes identity homomorphisms:

\[
\infer{\hat{\Gamma}\vdashh\Delta^{\mathsf{ID}}\,\hat{:\,(\gamma : \blc{\Delta^\A})\ra \blc{\Delta^\M}\,\gamma\,\gamma}}
      {\hat{\Gamma}\vdash\Delta}
\hspace{1em}
\infer{
  \hat{\Gamma}\vdashh A^{\mathsf{ID}}\,\hat{:\,(\gamma : \blc{\Delta^\A})(t : \blc{A^\A}\,\gamma)\ra \blc{A^\M}\,\gamma\,\gamma\,(\blc{\Delta^{\mathsf{ID}}}\,\gamma)\,t\,t }}
      {\hat{\Gamma}\semicol\Delta\vdash A}
\]

\[
\infer{
  \hat{\Gamma}\vdashh t^{\mathsf{ID}}\,\hat{:\, (\gamma : \blc{\Delta^\A})\ra
       \blc{t^\M}\,\gamma\,\gamma\,(\blc{\Delta^{\mathsf{ID}}}\,\gamma) = \blc{A^{\mathsf{ID}}}\,\gamma\,(\blc{t^\A}\,\gamma) }}
      {\hat{\Gamma}\semicol\Delta\vdash t : A}
\]

Here, the interpretation of terms witnesses functoriality: $t^\M$ maps
identity morphisms in $\Delta$ to displayed identity morphisms in $A$.
We translate $\U$ to identity functions:
\[
\U^{\mathsf{ID}}\,\hat{\gamma\,A} :\equiv \hat{\lambda\,(x : A).\, x}
\]
Note that $\U$ and $\U[\sigma]$ are definitionally equal in the theory of
signatures, where $\sigma$ is a parallel substitution. The translation of the
latter would be the following (omitting many details, including the handling of
substitutions in the translation):
\[
(\U[\sigma])^{\mathsf{ID}}\,\hat{\gamma\,A} \equiv \hat{\TR_{(\lambda\,x. A\ra A)}\,(\sigma^{\mathsf{ID}}\,\gamma)\,(\lambda\,x.\, x)}
\]
Here, $\hat{\sigma^{\mathsf{ID}}\,\gamma}$ yields an equation for functoriality of
$\hat{\sigma}$. Since the transport is constant, the result is propositionally
equal to $\hat{\lambda\,x.\, x}$. However, that is not enough, since we need to
preserve $\U = \U[\sigma]$ up to definitional equality.

There are numerous unavoidable instances of such failures of strict preservation
of definitional equalities for $\blank^{\mathsf{ID}}$, and they similarly arise when we
try to construct composition of homomorphisms. We consider three potential
solutions for this strictness problem:
\begin{enumerate}
\item Solving the coherence problem. This would allow us to interpret signatures
      into the metatheory, allowing preservation of definitional equality up
      to propositional metatheoretic equality.
\item Reformulating the syntax of signatures so that definitional equalities
      become weak propositional equalities. We already use weak $\beta$ for $\J$,
      can we do so elsewhere? However, this would result in an unusual and very
      inconvenient syntax of signatures, because we would need to weaken even
      basic substitution rules.
\item Instead of interpreting signatures into an UIP-free type theory, we
      interpret them into classical combinatorial structures for higher groupoids
      and categories, e.g.\ into simplicial sets. The drawback is that now we do not
      have the convenient synthetic notion of higher structures, provided by the syntax
      of type theory, and instead we have to manually build up these
      structures. Needless to say, this makes machine-checked formalization much
      more difficult. We have found mechanized formalization invaluable for the
      current paper, and it would be painful to abandon it in further research.
\end{enumerate}

@@

@@

There are three additional development artifacts to the current work: a Haskell
implementation, a shallow Agda formalization and a deep Agda formalization. All
three are available from \url{https://github.com/akaposi/hiit-signatures}.

The Haskell implementation takes as input a file which contains a of a
$\hat{\Gamma}\vdash\Delta$ signature. Then, it checks the input with respect to
the rules in Figure \ref{sigrules}, and outputs an Agda-checkable file which
contains algebras, homomorphisms, displayed algebras and sections for the
input signature.

It comes with examples, including the ones in this paper, the
inductive-inductive dense completion \cite[Appendix A.1.3]{forsberg-phd} and
several HITs from \cite{HoTTbook} including the definition of Cauchy reals. It
can be checked that our implementation computes the expected elimination
principles in these cases.

The shallow Agda formalization embeds both the source and target theories
shallowly into Agda: it represents types as Agda types, functions as Agda
functions, and so on. We also leave the $\blank^\A$ operation implicit. We state
each case of translations as Agda functions from all induction hypotheses to the
result type of the translation, which lets us ``typecheck'' the translation. We
have found that this style of formalization is conveniently light, but remains
detailed enough to be useful. We also generated some of the code of the Haskell
implementation from this formalization.

The deep Agda formalization deeply embeds the theory of signatures as a CwF with
additional structure, in the style of \cite{ttintt}. However, it embeds the
external type theory shallowly as Agda, and we model dependency on external
$\hat{\Gamma}$ contexts with Agda functions. We formalize $\blank^\A$,
$\blank^\D$, $\blank^\M$ and $\blank^\S$ translations as a single model
construction in Agda. This setup greatly simplifies formalization, since we do
not have to reason explicitly about definitional equality in the external
syntax, but we can still reason directly about preservation of definitional
equalities in the theory of signatures. This semi-deep formalization can be in
principle converted into a fully formal syntactic translation, because we prove
all preservations of definitional equalities by $\refl$. Due to technical
challenges, this formalization uses transport instead of $\J$ in the source
theory, but this still covers a rather large class of HIIT definitions.

%--------------------------------------------------------------------------------
